The Creation of a Programming Language

### 2022-10-08 paper.tex
continue writing on paper

% the name of the programming language
\newcommand{\name}{-the language- }

@@

\subsection{The Art of Abstraction}
When dealing with problems in general, a helpful method of going about can be
to create an abstraction. This is also the case when building a programming
language interpreter. Like that, an interpreter is often times subdivided into several
components, each solving a discrete problem. One way to define
such components could be the following:
\begin{enumerate}\label{simple interpreter}
    \item Create meaningful groups of characters from the input string. This
        process is called tokenizing and is done with the component called
        tokenizer or lexer. As an example, the string "aa+  bb" could result
        in three tokens, namely "aa", "+" and "bb", in case space characters
        are considered unimportant.
    \item Ordering the tokens into a so-called parse-tree or syntax-tree.
        Creating an order is important, as some calculations depend on others, e.g. multiplication
        is always done before addition. TODO:example
    \item Technically, the parse tree could now be interpreted directly, and
        an output could be produced, as the interpreter know knows what to do
        and in what order to perform computations.
\end{enumerate}
As seen above, these components, or stages as they will be called in this paper,
each input and output data, where the output of one stage is the input of the
respective next stage.

@@

Source code can be found on GitHub TODO: insert link

\subsubsection{The Entry Point}
The entry point of the project lies in the src/main.c file.
The file implements the C main function, which is the entry-point of any
C application in general. Since the file is so small, here is its source code:
\code{src/main.c}{1}{999}

@@

Here is a summary of what happens in the src/main.c file: the src/interpreter/interpreter.h
header file is included. In the main function, the program decides to either
read a program from the standard input or from a file. Then, the interpret function
from the said included header is called with the selected input.

@@

After that, variables with names are 'localized', meaning their names are
replaced by addresses where their value is stored by src/interpreter/localizer.c.
The advantage of performing this step so early is that mapping variable names
to memory is a rather resource intensive task with time complexity of maximally 
O(log n), n being the amount of different variables.

@@

\subsubsection{The Tokenizer}
The tokenizer is implemented in src/interpreter/tokenizer.c. Its job is to
take source code in plain-text format as input and output so-called tokens.
These tokens represent groups of characters that belong together. In this paper's
implementation, each token has a strict type and a text content.
The different types of tokens and other thins are defined in the header file
src/interpreter/tokenizer.h.

Here is an example of tokenization:
An input of 'std.test==23.4' would result in the following tokens according
to the tokenizer implementation of this paper:
\begin{enumerate}
    \item Token of type FIELD with content 'std.text'.
    \item Token of type SYMBOL with content '=='.
    \item Token of type NUMBER with content '23.4'.
\end{enumerate}
FIELD type tokens are later replaced by REFERENCES during the stage of localization
and SYMBOL type tokens are mapped to EXPR by src/implementation/mappings/mapop.c
during parsing. NUMBER type tokens are replaced by an actual number type, which
is C's 'long double' type as specified in src/interpreter/bytecode.h.

@@

The syntax tree in the implementation of \name consists of 'stnode' type
pointers, as defined in src/interpreter/parser.h. Each such node depending
on its 'nodetype', declared in src/interpreter/tokenizer.h,
is either a value-node storing a pointer to some data, e.g. a number or text, or a
parent-node referencing two child nodes and an operator, e.g. two numbers as
children and a +-operator to signify an addition.

The difficult part of implementing a parser is precedence. Presedence and
associativity define the order of execution of statements. A simple example
would be to perform multiplication before addition, e.g. "2+3*4" would
result in 14 and is not the same as "(2+3)*4". Therefore, each operator is
assigned a precedence or priority, as well as an associativity, which is
either 'Left to Right' or 'Right to Left'. An example of associativity is
'20 / 10 / 2'. Both in C as well as in \name, the division operator is
left-to-right associative, meaning the stated expression is the same as
'(20 / 10) / 2' and not as '20 / (10 / 2)'. Both precedence and associativity
of each operator are defined in the file mapop.gperf.

@@

Following is a piece of pseudo-code to better illustrate the inner workings of the
parser which is based on a Pratt-parser. 
This is only an example and not completely accurate to the actual parser used
by this project.
TODO: validate code
\code{pseudocode/parser.c}{1}{999}

\section{Results}
\subsection{Limitation}
Token input size etc.
Single threaded

@@

\subsection{Interesting Ideas untouched}
Creating the ideal programming language takes a lot of planning and sketching
out. This section of the paper gives insight on interesting ideas or concepts
that are used in other programming languages, not \name.

>>>>

% the name of the programming language
\newcommand{\name}{\emph{TestScript }}

@@

\subsection{The Art of Abstraction}
When dealing with problems in general, a helpful method of going about can be
to create an abstraction. This is also the case when building a programming
language interpreter. Like that, an interpreter is often times subdivided into several
components, each solving a discrete problem. One way to define
such components could be the following:
\begin{enumerate}\label{simple interpreter}
    \item Create meaningful groups of characters from the input string. This
        process is called tokenizing and is done with the component called
        tokenizer or lexer. As an example, the string 'aa+  bb' could result
        in three tokens, namely 'aa', '+' and 'bb', in case space characters
        are considered unimportant.
    \item Ordering the tokens into a so-called parse-tree or syntax-tree.
        Creating an order is important, as some calculations depend on others, e.g. multiplication
        is always done before addition. TODO:example
    \item Technically, the parse tree could now be interpreted directly, and
        an output could be produced, as the interpreter know knows what to do
        and in what order to perform computations.
\end{enumerate}
As seen above, these components, or stages as they will be called in this paper,
each input and output data, where the output of one stage is the input of the
respective next stage.

@@

\subsubsection{The Entry-Point}
The entry point of the project lies in the src/main.c file.
The file implements the C main function, which is the entry-point of any
C application in general. Since the file is so small, here is its source code:
\code{src/main.c}{1}{999}

@@

Here is a very quick summary of what happens in the src/main.c file: the src/interpreter/interpreter.h
header file is included. In the main function, the program decides to either
read a program from the standard input or from a file. Then, the interpret function
from the said included header is called with the selected input.

@@

After that, variables with names are 'localized', meaning their names are
replaced by addresses where their value is stored by src/interpreter/localizer.c.
The advantage of performing this step so early is that mapping variable names
to memory is a rather resource-intensive task with time complexity of maximally 
O(log n), n being the amount of different variables.

@@

\subsubsection{The Tokenizer}
The tokenizer is implemented in src/interpreter/tokenizer.c. Its job is to
take source code in plain-text format as input and output so-called tokens.
These tokens represent groups of characters that belong together. In this paper's
implementation, each token has a strict type and a text content.
The different types of tokens and other things are defined in the header file
src/interpreter/tokenizer.h.

Here is an example of tokenization:
An input of 'std.test==23.4' would result in the following tokens according
to the tokenizer implementation of this paper:
\begin{enumerate}
    \item Token of type FIELD with content 'std.text'.
    \item Token of type SYMBOL with content '=='.
    \item Token of type NUMBER with content '23.4'.
\end{enumerate}
FIELD type tokens are later replaced by REFERENCE tokens during the stage of localization
and SYMBOL type tokens are mapped to EXPR by src/implementation/mappings/mapop.c
during parsing. NUMBER type tokens are replaced by an actual number type, which
is C's 'long double' type as specified in src/interpreter/bytecode.h.

@@

The syntax tree in the implementation of \name consists of 'stnode' type
pointers, as defined in src/interpreter/parser.h. Each such node depending
on its 'nodetype', declared in src/interpreter/tokenizer.h,
is either a value-node storing a pointer to some data, e.g. a number or text, or a
parent-node referencing two child nodes and an operator, e.g. two numbers as
children and a plus-operator to signify an addition.

The difficult part of implementing a parser is precedence. Precedence and
associativity define the order of execution of statements. A simple example
would be to perform multiplication before addition, e.g. '2+3*4' would
result in 14 and is not the same as '(2+3)*4'. Therefore, each operator is
assigned a precedence, or priority, as well as an associativity, which is
either 'Left to Right' or 'Right to Left'. An example of associativity is
\emph{'20 / 10 / 2'}. Both in C as well as in \name, the division operator is
left-to-right associative, meaning the stated expression is the same as
\emph{'(20 / 10) / 2'} and not as \emph{'20 / (10 / 2)'}. Both precedence 
and associativity are properties that are defined in the file mapop.gpref for
each operator.

@@

Following is a piece of pseudocode to better illustrate the inner workings of the
parser which is based on a Pratt-parser. 
This is only an example and not completely accurate to the actual parser used
by this project.
TODO: validate code
\code{pseudocode/parser.c}{1}{999}

\subsubsection{The Bytecode Compiler}
The bytecode compiler is part of the interpreter. Its job is to create a list
of byte-sized instructions and data from the syntax-tree the parser created.
Having such an intermediary bytecode format, instead of directly interpreting
the parse-tree can be beneficial. This is the case when the same piece of code
has to be run several times. Instead of generating a time and memory overhead 
having to walk along tree-nodes recursively, only the list of instructions in
the bytecode can be traversed, which is much faster.

Not to be confused with machine-native bytecode, the bytecode compiler in \name
generates byte-sequences in an internal format, that are then interpreted. A
JIT compiler would instead compile the AST down to machine-native bytecode, which
could run directly on the system's hardware and would not have to be interpreted.

To generate such a sequence of bytes, \name recursively reads through the
binary tree created by the parser and pushes found instructions onto a stack,
which is in this case used similarly to a variable-sized list.
The algorithm used is inspired by Edsger Dijkstra's 'Shunting yard 
algorithm'\footcite{shunting_yard}, traversing the tree post-order to create
a prefix notation.

A prefix notation is a way to write down mathematical expressions, except
operators precede their operands, in contrast to the more common infix notation,
where operands surround the operator.

In src/interpreter/bytecode.c we can see, that first, the left operand and then
the right operand are added onto the stack of bytecode. Therefore, when
popping from the stack and evaluating instructions, the operands will have to
be read in reverse.
Here is an example:
\begin{enumerate}
    \item Assume the following mathematical expression in infix notation:
        \begin{center}
            \emph{'1 - 2 * 4 / (5 + 6)'}
        \end{center}
    \item In common prefix notation, also called Polish notation, the
        operators would now precede the operands as such:
        \begin{center}
            \emph{'- 1 (/ (* 2 4) (+ 5 6))'}
        \end{center}
    \item But in \name the operands are pushed onto the stack is reverse order.
        This would look like the following:
        \begin{center}
            \emph{'- (/ (+ 6 5) (* 4 2)) 1'}
        \end{center}
    \item The brackets are not needed, because for every operator, the amount
        of operands is known. This is the case as e.g. the subtraction operator
        and the negative operator, which are both a minus sign, are distinguished
        beforehand. This means the above statement would look like this:
        \begin{center}
            \emph{'- / + 6 5 * 4 2 1'}
        \end{center} 
        This is an ideal abstraction of how data 
        is represented internally in \name.
\end{enumerate}

\subsubsection{The Runtime}
This is the last stage of interpreting program code as implemented in \name.
The runtime steps through the intermediary bytecode created by the bytecode
compiler and evaluates expressions by plugging in the provided values.
All available expressions are implemented in 
src/interpreter/processing/implementations.c.

The evaluation of expressions also works in a recursive manner. Here is an 
example of \name's bytecode notation and how it would be evaluated at runtime:
\begin{itemize}
    \item Assume the input in \name's bytecode to resemble the following expression: 
        \begin{center}
            \emph{'- 6 * 3 4'}
        \end{center}
        In infix notation, this would be equivalent to 
        \begin{center}
            \emph{'(4 * 3) - 6'}.
        \end{center}
    \item The runtime will read through the bytecode left-to-right. First it will
        notice the subtraction-operator and accordingly assume that two
        operands must follow.
    \item The processing function will call itsself
        and notice the 6, a number, and therefore a value. The function simply returns the 6
        to the subtraction expression. This 6 will be the right-hand operator
        of the subtraction.
    \item The subtraction will then call the processing function again to find
        the left operand. As the processing function reads a multiplication operator,
        it now recursively reads through that expression's right and then left
        operator, which would be 3 and for respectively and compute and return
        the result of the multiplication, which is 12.
    \item Now the subtraction knows its right- and left-hand operators, 6 and 12,
        and again, calculate and return the result, which is 6.
\end{itemize}

That's it! With all the possible expressions available, entire programs
can now be interpreted without a problem.

\section{Results}
\subsection{Limitation}
\name is a programming language without a doubt. But it is not perfect in any
way. There are limitations that make it practically unusable for actual
professional projects. In this section, a few shortcomings of \name are listed
and reasoned over.

\subsubsection{Memory Leaks}
Memory leaks happen, when a program reserves memory to store some data, but
never frees up the space for use for other programs, even though the data
has long been discarded and is of no further use.
In \name, memory leaks are a major problem. Even though barely noticable for small
programs it interprets, \name does not have proper setups in place to free up
all used memory that is not used any longer.

Other programming languages, notably object-oriented languages such as Java,
use Garbage-Collection, which automatically tracks unused allocated memory and
frees it up automaticadlly. While this is also the case with \name, there are
many internal flaws that have nothing to do with variables used in a \name
program, but rather internally.

A major reason for the memory leaks, but not an excuse, is the nature of the
C programming language the \name interpreter was written in. The C language and
compiler do not keep track of states in which certain blocks of memory are
forgotten about and warn the developer, instead, memory leaks happen.
More modern C alternatives such as Rust have a ownership-system in place,
in which any allocated memory has a function as an owner, and when the function
is done processing, the memory is automatically freed. Such systems are a
helpful tool for developers to create better programs with low mental overhead,
and are what programming language design should be all about.

\subsubsection{Single-Threaded}
The \name interpreter does not implement any multi-threaded, meaning parallel
running, programs as most other languages do. Instead, any \name program can
only be interpreted and run sequentially. Since most modern computer processors
are equipped with a handful of processing cores and threads, not being able to
make use of those is a major downside and in itsself a reason not to use this
language.

\subsubsection{Token input size}
In the current iteration, each token's content's size is limited to 256 bytes,
which makes it impossible to use e.g. variable names longer thatn 256 ASCII
characters. This would probably be the easyest thing to fix of here-listed 
problems of \name, but since 256 characters are long enough for 
illustration purposes, fixing this is not worth spending time on.

\subsubsection{Lack of a Development Environment}
This is certainly the greatest problem of \name, aside the countless smaller
bothering aspects.

Pretty much any good, modern programming language does not only consist
of it's specification and a compiler or interpreter, but of many more
tools surrounding it.
Such tools often include a package-manager, with which people around the world
can share code. Examples would be npm\footcite{npm} or cargo\footcite{cargo}, 
the 'Node Package Manager' for JavaScript applications and and Rust's package 
manager respectively.

@@

\subsection{Interesting Ideas untouched}
Creating the ideal programming language takes a lot of planning and sketching
out. This section of the paper gives insight on interesting ideas or concepts
that are great to have and are used in other programming languages, but not 
could not be implemented into \name.
